Comparing Power Optimizations in the Reorder Buffer

### 2011-06-14 paper/bg.tex
Dist ROB section in paper done.
Editted.

\section{Background}
    An out of order execution system has two additional stages 
to the lifetime of the instruction on top of the traditional pipeline 
stages.  Each of which a small memory array to keep track of instruction flow.
The first addition, the instruction queue(IQ) which holds instructions that are 
waiting to be executed.  Each entry will hold the instruction -- including 
type, operand registers, and destination registers.  The second structure 
holds the instructions that are in flight but not yet committed to the register
file. Each entry will hold both PC of the instruction, as well as the destination
register and the result of the execution unit.  There is also a flag that will
mark the result of the entry as valid.

Replacing the traditional instruction dispatch stage, An placeholder will be made for the
instruction in ReOrder Buffer (ROB), and the instructions will be placed
in the instruction queue(IQ). In this stage instructions will first wait for their operands if not already
available.  When all operands are available, the instruction is ready to
be executed so wait until its appropriate execution unit is free.  At this
point the instruction exits the IQ and begins execution.

Upon execution completion, the output value is stored in the ROB with its 
corresponding entry and that entry is marked as valid or completed with a 
1 bit flag.  An entry is committed from the ROB to the architectural register
file (ARF) when that entry is the oldest (lowest PC) and its result has been 
marked as valid. 
\subsection{Circular ROB}
The circular ROB is one of the most basic implementations of the ROB.  Simply using
a circular buffer as the structure to hold each entry.  Since the ROB is basically
just memory, there are two special slots for to keep track of the head and tail
pointer to create the circular buffer.  As entries are written into the ROB, the
entry is written at the tail pointer and then the tail pointer is incremented.
Likewise when an entries is written to the ARF, the head pointer is incremented.
The circular buffer method is widely used since it inherently keeps track of the 
oldest instruction (which is at the head pointer), as well as which entry slots
are used or empty.   However it may not be very power efficient if there is a large
number of empty slots at any given point of the execution.    
\subsection{Dynamic ROB}

>>>>

\section{Background}
    An out of order execution system has two additional stages 
to the lifetime of the instruction on top of the traditional pipeline 
stages.  Each of which is a small memory array to keep track of instruction flow.
The first addition, the instruction queue(IQ), holds instructions that are 
waiting to be executed.  Each entry will hold the instruction -- including 
type, operand registers, and destination registers.  The second structure 
(ROB) holds the instructions that are in flight but not yet committed to the register
file. Each entry will hold both PC of the instruction, as well as the destination
register and the result of the execution unit.  There is also a flag that will
mark the result of the entry as valid.

Replacing the traditional instruction dispatch stage, a placeholder will be made for the
instruction in reorder buffer and the instructions will be placed
in the instruction queue(IQ). In this stage instructions will first wait for their operands if not already
available.  When all operands are available, the instruction is ready to
be executed and waits until its appropriate execution unit is free.  At this
point the instruction exits the IQ and begins execution.

Upon execution completion, the output value is stored in the ROB with its 
corresponding entry and that entry is marked as valid or completed with a 
1 bit flag.  An entry is committed from the ROB to the architectural register
file (ARF) when that entry is the oldest (lowest PC) and its result has been 
marked as valid. 
\subsection{Circular ROB}
The circular ROB is one of the most basic implementations of the ROB.  It is implemented simply using
a circular buffer as the structure to hold each ROB entry.  Since the ROB is basically
just memory, there are two special slots to keep track of the head and tail
pointers of the circular buffer.  As entries are written into the ROB, the
entry is written at the tail pointer and then the tail pointer is incremented.
Likewise when a entry is written to the ARF, the head pointer is incremented.
The circular buffer method is widely used since it inherently keeps track of the 
oldest instruction (which is at the head pointer), as well as which entry slots
are used or empty.   However it may not be very power efficient if there is a large
number of empty slots at any given point of the execution.

\subsection{Dynamic ROB}

### 2011-06-14 paper/proj.tex
Dynamic buffer results write up.
Editted.

For the purposes of our power comparisons, each of the above instruction workloads are normalized to generate 10000 instructions.  Such a large instruction window is ideal to allow buffers to fill and behavioral characteristics to develop.  The specific values used for each workload are listed in Table ~\ref{tab:insprofile}.
\subsection{Model Implementation}
The following section details the model implementations in our simulator.  The basic re-order buffer model we use as a base-line for comparison is a standard circular ROB model.  Section ~\ref{sec:circ} discusses its implementation.  We further model three optimizations for re-order buffers in an attempt to improve overall power usage: 1) Distributed (Cluster based) ROB, 2) Dynamically Sized ROB, 3) Reduced Output Port ROB (Retention Latch).  Each of these optimized models are built on top of the underlying circular ROB model to provide as close a behavioral and power comparison as possible.  Each of the optimizations, as well as the standard circular reorder buffer, are described in detail below.
\subsubsection{Circular ROB}
\label{sec:circ}
The circular ROB is the basic current implementation of the reorder buffer, so far each of the other ROB designs will use the 
circular array for each instruction entry.  Each instruction entry will have a flag to determine if its result is valid, a 
result field (currently 32 bits), the corresponding register id tag and the PC of the instructions.  There are two additional
fields in each entry that is an artifact of our simulation, the first of which is a flag to denote a floating point entry, and 
mark that the result of the next entry is a continuation of the current one.  The second element that is an artifact of the
simulator is the cycles counter, which acts as a timing simulator and will determine when the result will finally be produced
by the execution unit.  The ROB is simply an array of these entry structures, and two pointers (32 bits) that will point
to the head or tail of the circular FIFO buffer.  Also the variables used for power tabulation are a part of the structure as 
another artifact of our simulator.

@@

\begin{table*}[!tb]
\centering
\caption{FP Heavy Instruction Profile Results(B1)}
\begin{tabular}{|c|c|c|c|c|c|} \hline
    & Circular  & Distrubuted & Dynamic & Dynamic (optimized) & OOP\\ \hline
Cycles&82,722	&50,541&    	84,211	&84,078&        	82,153\\ \hline
Bit Cycles&1,483,701,792&	1,399,581,372&	1,453,144,453	&1,470,555,825	&1,522,130,784\\ \hline
Trans. to High&1,967,483&	1,719,957	&1,942,948	&1,930,476	&3,235,293\\ \hline
Reg Comp&6,725,088&	6,631,913&	6,710,777&	6,718,043	&11,160,291\\ \hline
\end{tabular}
\end{table*}

\begin{table*}[!tb]
\centering
\caption{int Heavy Instruction Profile Results(B2)}
\begin{tabular}{|c|c|c|c|c|c|} \hline
    & Circular  & Distrubuted & Dynamic & Dynamic (optimized) & OOP\\ \hline
Cycles&25,781&	25,794&	26,079	&25,783	&25,780\\ \hline
Bit Cycles&462,408,016	&714,287,448	&310,408,263	&317,109,601	&477,651,840\\ \hline
Trans. to High&1,347,679&	1,134,479	&1,363,418&	1,369,396	&2,262,036\\ \hline
Reg Comp&3,220,687	&3,221,904	&3,219,866	&3,221,335&	5,155,838\\ \hline
\end{tabular}
\end{table*}

@@

Reg Comp&1,688,252&	1,436,544	&1,655,666	&1,657,563	&2,802,135\\ \hline
\end{tabular}
\end{table*}

\subsubsection{Omitted Output Port ROB (Retention Latch)}
The main difference between the Circular ROB and the Omitted Output Port (OOP) ROB is that the OOP adds in a retention latches and is not allowed to forward any results directly to DU.  This our implementation of the OOP ROB simply extends the circular ROB, adding in a structure for the retention latches.  The latches is modeled very similar to the ROB entries, an array
of a latch entry structure.  Each latch entry contains a 32 bit value for the results to be stored as well small tag to the corresponding destination register of the result.  As an artifact of our simulation a flag to denote that the entry is part of a floating point operation is required as well.  Also an additional counter was added to count the additional re-forwarding when a value is committed to the ARF but also needs to be forwarded to the DU unit as well.  Though this is not an increasing number of ports driven, it may affect future power calculations relating to capacitance.  Lastly as changing the semantic of the number of forwards to the DU, the count refers to the number time the entry is read from a retention latch.

As a few minor changes to the behavioral function of the OOP ROB, the first of which is the lack of the ability to forward operands.  Thus in the simulated stage of reading from the instruction stream, when we would normally check if the operand is in the ROB and forward.  We instead do the following, we still check to see if the operand is in the ROB, and the retention latches. If the entry is not in the ROB, or the latches, the operand would be available in the ARF or memory, in the converse case,  the entry in both, the ROB, and the latches, the operand is forwarded from the latches and the counter is incremented .  In the case that the entry exists in the ROB, but not in the latches, a flag would be set in the ROB to forward the result when committing to the ARF. The last case, that is the entry exists in the latches and not in the ROB, means that the result has already been committed and the result could be forwarded from the ARF.  Next in our update entries (when the cycle counter expires -- execution has completed) we would add in the new entry to the retention latch.  Lastly when the ROB entires are valid, and ready to be committed to the ARF,
if the forwarding flag is set, the forwarding counter is incremented.

\subsection{Power Tabulation Implementation}
In order for our simulation to determine the power utilization, we do not try and equate each event with an associated cost in joules.  As technology changes, and new techniques are developed to decease circuit level power usages, our simulation will become less and less accurate.  Instead we chose to simply tabulate each event that is associated with power cost, such as bit flips from 0 to 1, register comparator usage, and bit lines driven.  To help measure leakage power and overall states of the ROB, we also keep track of the time (in cycles) each bit stays in the high state, or the low state.

>>>>

For the purposes of our power comparisons, each of the above instruction workloads are normalized to generate 10000 instructions.  Such a large instruction window is ideal to allow buffers to fill and behavioral characteristics to develop.  The specific values used for each workload are listed in Table ~\ref{tab:insprofile}.
\subsection{Model Implementation}
The following section details the model implementations in our simulator.  The basic re-order buffer model we use as a base-line for comparison is a standard circular ROB model.  Section ~\ref{sec:circ} discusses its implementation.  We further model three optimizations for re-order buffers in an attempt to improve overall power usage: 1) Distributed (Cluster based) ROB, 2) Dynamically Sized ROB, 3) Reduced Output Port ROB (Retention Latch).  Each of these optimized models are built on top of the underlying circular ROB model to provide as close a behavioral and power comparison as possible.  Each of the optimizations, as well as the standard circular reorder buffer, are described in detail below.
\subsubsection{Circular ROB}
\label{sec:circ}
The circular ROB is the basic current implementation of the reorder buffer -- so far each of the other ROB designs will use the 
circular array for each instruction entry.  Each instruction entry will have a flag to determine if its result is valid, a 
result field (currently 32 bits), the corresponding register id tag and the PC of the instructions.  There are two additional
fields in each entry that is an artifact of our simulation, the first of which is a flag to denote a floating point entry, and 
mark that the result of the next entry is a continuation of the current one.  The second element that is an artifact of the
simulator is the cycles counter, which acts as a timing simulator and will determine when the result will finally be produced
by the execution unit.  The ROB is simply an array of these entry structures, and two pointers (32 bits) that will point
to the head or tail of the circular FIFO buffer.  Also the variables used for power tabulation are a part of the structure as 
another artifact of our simulator.

@@

\begin{table*}[!tb]
\centering
\caption{FP Heavy Instruction Profile Results(B1)}
\begin{tabular}{|c|c|c|c|c|c|} \hline
    & Circular  & Distrubuted & Dynamic & Dynamic (optimized) & OOP\\ \hline
Cycles&82,722	&50,541&    	84,211	&84,078&        	82,153\\ \hline
Bit Cycles&1,483,701,792&	1,399,581,372&	1,453,144,453	&1,470,555,825	&1,522,130,784\\ \hline
Trans. to High&1,967,483&	1,719,957	&1,942,948	&1,930,476	&3,235,293\\ \hline
Reg Comp&6,725,088&	6,631,913&	6,710,777&	6,718,043	&11,160,291\\ \hline
\end{tabular}
\label{tab:bench1}
\end{table*}

\begin{table*}[!tb]
\centering
\caption{int Heavy Instruction Profile Results(B2)}
\begin{tabular}{|c|c|c|c|c|c|} \hline
    & Circular  & Distrubuted & Dynamic & Dynamic (optimized) & OOP\\ \hline
Cycles&25,781&	25,794&	26,079	&25,783	&25,780\\ \hline
Bit Cycles&462,408,016	&714,287,448	&310,408,263	&317,109,601	&477,651,840\\ \hline
Trans. to High&1,347,679&	1,134,479	&1,363,418&	1,369,396	&2,262,036\\ \hline
Reg Comp&3,220,687	&3,221,904	&3,219,866	&3,221,335&	5,155,838\\ \hline
\end{tabular}
\label{tab:bench2}
\end{table*}

@@

Reg Comp&1,688,252&	1,436,544	&1,655,666	&1,657,563	&2,802,135\\ \hline
\end{tabular}
\label{tab:bench3}
\end{table*}

\subsubsection{Omitted Output Port ROB (Retention Latch)}
The main difference between the Circular ROB and the Omitted Output Port (OOP) ROB is that the OOP adds in retention latches and is not allowed to forward any results directly to DU.  Our implementation of the OOP ROB simply extends the circular ROB, adding in a structure for the retention latches.  The latches are modeled very similarly to the ROB entries as simple arrays
of a latch entry structure.  Each latch entry contains a 32 bit value for the results to be stored as well as a small tag for the corresponding destination register of the result.  As an artifact of our simulation a flag is stored to denote that the entry is part of a floating point operation.  Also an additional counter was added to count the occurrences of re-forwarding when a value is committed to the ARF but also needs to be forwarded to the DU unit as well.  Though this is not increasing number of ports driven, it may affect future power calculations relating to capacitance.  Lastly as changing the semantics of the number of forwards to the DU, the count refers to the number of times the entry is read from a retention latch.

There are a few minor changes to the behavioral function of the OOP ROB, the first of which is the lack of the ability to forward operands.  Thus in the simulated stage of reading from the instruction stream, we would normally check if the operand is in the ROB and forward.  We instead do the following: we still check to see if the operand is in the ROB but also check the retention latches. If the entry is not in the ROB, or the latches, the operand would be available in the ARF or memory.  In the converse case,  the entry is in both the ROB and the latches, the operand is forwarded from the latches and the counter is incremented .  In the case that the entry exists in the ROB, but not in the latches, a flag would be set in the ROB to forward the result when committing to the ARF. The last case, that is the entry exists in the latches and not in the ROB, means that the result has already been committed and the result could be forwarded from the ARF.  Next in our update entries (when the cycle counter expires -- execution has completed) we would add in the new entry to the retention latch.  Lastly when the ROB entires are valid, they are ready to be committed to the ARF,
if the forwarding flag is set, the forwarding counter is incremented.

\subsection{Power Tabulation Implementation}
In order for our simulation to determine the power utilization, we do not try and equate each event with an associated cost in joules.  As technology changes, and new techniques are developed to decease circuit level power usages, our simulation will become less and less accurate.  Instead we chose to simply tabulate each behavioral event that is associated with power cost, such as bit flips from 0 to 1, register comparator usage, and bit lines driven.  To help measure leakage power and overall states of the ROB, we also keep track of the time (in cycles) each bit stays in the high state, or the low state.

### 2011-06-14 paper/start.tex
Editted.

\begin{abstract}
        In an effort to increase instruction level parallelism techniques such
as pipelining, out of order execution and operand forwarding.  However
these efforts come at a large power cost.  Out of order execution requires
a ReOrder Buffer (ROB), to commit the registers in program order.  This
structure has grown to be monolithic with numerous entries and read/write ports.
There is no current methods to efficiently and equally compare various
ROB advancements to improve power efficiency. We propose a loosely cycle
accurate model that keeps track of the power usages of the ROB.  Using the 
simulation tool, we are able to compare and contrast the affects of improvements
made on the ROB.  This also allows for a first line test to validate proof of
concepts for future power efficient ROB designs.
\end{abstract}

@@

\section{Introduction}
Out of order execution is a common technique used to aggressively push instruction
throughput to meet the demands of modern computer users.  As the name suggests
this allows for instructions to be inflight while previous instructions may still be waiting 
for an available execution unit or operands.  To accomplish this, two additional structures/stages
are necessary  on top of the traditional pipelined architecture.  Both of these additional
structures are essentially memory elements. The first one to keep track of which instructions
are ready to be executed and the other known as the ROB, allows the processor  to commit the
results of execution in program order to maintain program correctness.

Through out its evolution the the ROB has grown to a large monolithic structure that consumes
on average 7\% of the overall processor power budget\cite{rabey}.  Various new implementations
of the ROB have been proposed in an effort to reduce power consumption.  Some of which 
dynamically change the the size of the ROB, or a banked buffer structure and other proposals argues in the 
extreme case to eliminate read ports all together\cite{kucuk}\cite{kucuk2} \cite{kucuk3}.

With the rapid change of manufacturing technologies and other impacts on energy utilization it 
becomes hard to determine the gain of each new proposed implementation.  We propose a new
a loosely cycle accurate ROB simulator that provides a fair comparison of  different ROB designs.
ROB\_pwr is a simulator that tabulates each action that would cost power.  This allows for a fair
comparison between each ROB design, without the concern for the underlying technology.

>>>>

\begin{abstract}
        In an effort to increase instruction level parallelism, techniques have been developed such
as pipelining, out of order execution and operand forwarding.  However
these efforts come at a large power cost.  Out of order execution requires
a ReOrder Buffer (ROB), to commit the registers in program order.  This
structure has grown to be monolithic with numerous entries and read/write ports.
There are no current methods to efficiently and equally compare various
ROB advancements to improve power efficiency. We propose a loosely cycle
accurate model that keeps track of the power usages of the ROB.  Using the 
simulation tool, we are able to compare and contrast the affects of improvements
made on the ROB.  This also allows for a first line test to validate proof of
concepts for future power efficient ROB designs.
\end{abstract}

@@

\section{Introduction}
Out of order execution is a common technique used to aggressively push instruction
throughput to meet the demands of modern computer users.  As the name suggests
this allows for instructions to be inflight while previous instructions may still be waiting 
for an available execution unit or operands.  To accomplish this, two additional structures/stages
are necessary  on top of the traditional pipelined architecture.  Both of these additional
structures are essentially memory elements. The first structure is to keep track of which instructions
are ready to be executed and the other, known as the reorder buffer (ROB), allows the processor to 
execute instructions out of order but commit the results of execution in program order to maintain 
program correctness.

Throughout its evolution the the ROB has grown to a large monolithic structure that consumes
on average 7\% of the overall processor power budget\cite{rabaey}.  Various new implementations
of the ROB have been proposed in an effort to reduce power consumption.  Some of which 
dynamically change the the size of the ROB or propose a banked buffer structure and other proposals argues in the 
extreme case to eliminate read ports all together\cite{kucuk}\cite{kucuk2} \cite{kucuk3}.

With the rapid change of manufacturing technologies and other impacts on energy utilization it 
becomes hard to determine the gain of each new proposed implementation.  We propose a new
and loosely cycle accurate ROB simulator that provides a fair comparison of  different ROB designs.
ROB\_pwr is a simulator that tabulates each hardware action that would cost power.  This allows for a fair
comparison between each ROB design, without the concern for the underlying technology.
